\textbf{ Spectrophotometric parallaxes with linear models:\\ Accurate distances for luminous red-giant stars }

### 2018-10-15 paper/sp.tex
more to-do items

% To-do items
% -----------
% - Coryn's point that we should carefully explain how our model is geometric.
% - check with WISE experts that we have the correct WISE photometry
% - write words about other linear models we have made or could make, probably in the discusssion.
% - are we citing all the proper DR2 papers we need to cite?
% - HWR: Footnote to explain the use of the word ``feature'' for astronomers.
% - HWR: Figure 3: Note that the y-axis is positive-definite!
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email.
% - add paragraph about AGB stars and the color-magnitude diagram in the discussion.
% - Cite and describe Eilers et al.
% - comment on parallax offset being a function of position and color
% - remove Cannon plot and all references for the moment
% - comment on new plot of new precision on parallax compared to Gaia's precision
% - write caption and paragraph about the new figure showing how much better we are than Gaia at large distance
% - need a snippet for Table 1.

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not highly informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia--\zmass--\wise\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$.
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets.
We can now get 10-percent distance estimates out to heliocentric distances of 20\,kpc,
and make global maps of the Milky Way's disk.
\end{abstract}

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
\begin{description}
\item[features] Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ (\citealt{apogee}) spectrum,
plus the $G$, $\BP$, and $\RP$ photometry from \gaia\ (\citealt{gaia}),
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise\ (\citealt{wise}).
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation is shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of less than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Parent Sample, the middle panel shows the Training Set
  stars (which pass \gaia\ quality checks; see text), and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. The fractional precision of the prediction in the
  right panel is better than 9 percent.\label{fig:xval}}
\end{figure}

@@

This 9-percent precision estimate is conservative in the sense that it does not
deconvolve or correct for the contributions from the \gaia\ noise.
However it is also not fully conservative, in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction validation colored by relevant features.
  Each panel shows the predicted parallax \foreign{vs} the
  spectrophotometric parallax colored by a different
  data feature that might be relevant to the residual.\label{fig:residuals}}
\end{figure}

>>>>

% To-do items
% -----------
% - need a snippet for Table 1.
% - check with WISE experts that we have the correct WISE photometry
% - HWR: Figure 7: Caption needs to say what arrows represent.
% - Make sure we incorporate / respond to Ness comments from Aug 13 email.
% - Cite and describe Eilers et al.

@@

\begin{abstract}\noindent
% context
With contemporary infrared spectroscopic surveys like \apogee,
red-giant stars can be observed to distances and extinctions
at which \gaia\ parallaxes are not highly informative.
Yet the combination of effective temperature, surface gravity, composition, and age---all
accessible through spectroscopy---determines a giant's luminosity.
Therefore spectroscopy plus photometry should enable 
precise spectrophotometric distance estimates.
% aims
Here we use the \apogee--\gaia--\zmass--\wise\ overlap to train a data-driven model
to predict parallaxes for red-giant branch stars with $0<\logg\leq2.2$.
% methods
We employ (the exponentiation of)
a linear function of \apogee\ spectral pixel intensities and multi-band photometry
to predict parallax spectrophotometrically.
The model training involves no logarithms or inverses of the \gaia\ parallaxes,
and needs no cut on the \gaia\ parallax signal-to-noise ratio.
It includes an L1 regularization to zero out the contributions of
uninformative pixels.
The training is performed
with leave-out subsamples such that no star's astrometry is used even indirectly in its
spectrophotometric parallax estimate.
% results
The model implicitly performs a reddening and extinction correction in its parallax prediction,
without any explicit dust model.
We assign to each star in the sample a new spectrophotometric parallax estimate;
these parallaxes have uncertainties of a few to 15 percent, depending on data quality,
which is more precise than the \gaia\ parallax for the vast majority of targets, and
certainly any stars more than a few kpc distance.
We obtain 10-percent distance estimates out to heliocentric distances of 20\,kpc,
and make global maps of the Milky Way's disk.
\end{abstract}

@@

Our position is that a methodological technique is correct inasmuch as
it delivers correct or best results \emph{under its particular assumptions}.
In that spirit, we present here the assumptions of the method
explicitly.
\begin{description}
\item[features]\footnote{Here and everywhere in this \documentname, we
  use the word ``feature'' to refer to an input or predictor or
  independent variable for the regression.
  That is, our parallax predictions will be linear combinations of features.
  This is not common terminology in astrophysics, but it is standard in the machine-learning literature.}
Perhaps our most fundamental assumption is that the parallax
of a star can be predicted from the features we provide, which are
the full set of (pseudo-continuum-normalized) pixels from the \apogee\ (\citealt{apogee}) spectrum,
plus the $G$, $\BP$, and $\RP$ photometry from \gaia\ (\citealt{gaia}),
plus the $J$, $H$, and $K_s$ photometry from \zmass\ (\citealt{zmass}),
plus the $W_1$ and $W_2$ photometry from \wise\ (\citealt{wise}).
That is, we assume that these spectrophotometric \emph{features} are sufficient
to predict the parallax in the face of variations in stellar
age, evolutionary phase, composition, and other parameters, and also interstellar
extinction.

@@

We optimize two models, one for Training Set A and one for Training Set B.
The two models are applied to the two splits of the Parent Sample, Parent Sample
A and Parent Sample B, using the A-trained model on Parent Sample B and
\foreign{vice versa}.
For the purposes of assessing the accuracy and precision of the model, this
train-and-test framework constitutes a two-fold cross-validation.
Results of this cross-validation is shown in \figurename~\ref{fig:xval}.
For the stars with highest \gaia-measured parallax signal-to-noise,
the spectrophotometric model is predicting the \gaia\ parallaxes with little bias
and a scatter of less than 9~percent.
\begin{figure}
\includegraphics[width=\textwidth]{residuals.pdf}
\caption{The predictive accuracy of the model, in the two-fold cross-validation.
  Each panel shows \gaia\ astrometric parallaxes $\gparallax_n$ on the vertical axis
  and our spectrophotometric parallaxes $\sparallax$ on the horizontal axis.
  The left panel shows the full Parent Sample, the middle panel shows the Training Set
  stars (which pass \gaia\ quality checks; see text), and the right panel shows a
  subset of stars with very high astrometric parallax signal-to-noise. This latter
  set plays no particular role in the method, but it can be used to demonstrate or
  assess the prediction precision. The fractional precision of the prediction in the
  right panel is better than 9 percent. Note that by construction, there can be no negative
  spectrophotometric parallaxes; that is, the quantity plotted on the vertical axis is
  positive-definite.\label{fig:xval}}
\end{figure}

@@

This 9-percent precision estimate is conservative in the sense that it does not
deconvolve or correct for the contributions from the \gaia\ noise.
However it is also not fully conservative, in that the test is performed with
\gaia's best stars, which are bright, nearby, in low extinction regions, and
(because of these things) near Solar metallicity.
There could in principle be additional bias or scatter for stars that are in dustier
regions or at lower metallicities.
\figurename~\ref{fig:residuals} shows that
there is no suggestion of any such trends when we color the residuals by metallicity
or $H-W_2$ color (which is a reddening proxy).
However, that figure does show that stars with greater reddening do appear to have
larger scatter against the \gaia\ parallaxes.
Additional tests confirm that our spectrophotometric parallaxes have largest scatter
in the dustiest and most crowded regions, as expected given the relatively low angular resolution
photometry.
\begin{figure}
\includegraphics[width=\textwidth]{./residuals_training.pdf}
\caption{Prediction validation colored by relevant features.
  Each panel shows the predicted parallax \foreign{vs} the
  spectrophotometric parallax colored by a different
  data feature that might be relevant to the residual.
  Again the quantity plotted on vertical axis is positive-definite.\label{fig:residuals}}
\end{figure}
